title: "PCA - Cereals Analysis"
author: "Vivek Teja"
date: "6/19/2020"
output: html_document
---
Objective:we will first clean a dataset and create some simple visualizations. Then, we will
conduct a principal component analysis to reduce dimensionality. We use R to
answer the questions. The data were collected on the nutritional information and consumer rating of 77
breakfast cereals. For each cereal we have included 13 numerical variables and 2 categorical variables
(i.e. mfr and type).
Variable
Name
Description
Name: Name of cereal
mfr: Manufacturer of cereal. {A=American Home Food Products, G=General Mills,
K=Kelloggs, N=Nabisco, P=Post, Q=Qauker Oats, R=Ralston Purina}
type: {cold, hot}
calories: Calories per serving
protein: Grams of protein
fat: Grams of fat
sodium: Milligrams of sodium
fiber: Grams of dietary fiber
carbo: Grams of complex carbohydrates
sugars: Grams of sugars
potassL Milligrams of potassium
vitaminsL Vitamins and minerals-0, 25, or 100, indicating the typical percentage of FDA
recommended
shelf Display shelf: 1, 2, or 3, counting from the floor
weight: Weight in ounces of one serving
cups: Number of cups in one serving
rating: A rating of the cereals calculated by Consumer Reports

---
Loading the Cereals Data Set

```{r}
setwd("C:/Users/vamsi/Downloads")
library(readr)
library(ggplot2)
Cereals <- read_csv("Cereals.csv")
View(Cereals)
```
1. Regarding two categorical variables (mfr and type), use a single proper visualization to show the
distributions of both variables

```{r}
ggplot(Cereals, aes(type, ..count..)) + geom_bar(aes(fill = mfr), position = "dodge")
```

2. Calculate the average consumer rating for each combination of manufacturer (mfr) and type. Obtain a
data frame (table) that follows the structure below. Use “0” to represent missing value if there is no any
cereal for a certain combination of manufacturer and type. Take a screenshot of the data frame.
```{r}
library(sqldf)
library(tidyr)
library(reshape)
q2<- sqldf("select type,mfr,avg(rating) as avg from Cereals group by type,mfr")

a<- sqldf("select distinct type from Cereals")

b<- sqldf("select distinct mfr from Cereals")

c<-crossing(a, b)

q2_<-sqldf("select b.type, b.mfr, case when avg is null then 0 else avg end as avg from (select c.*, a.avg from c left join q2 a on c.type=a.type and c.mfr =a.mfr)b ")

v<-data.frame(cast(q2_, type ~ mfr))
v
```
3) Use a proper visualization to compare the distribution of calories by hot vs. cold cereals. What does
this plot show us regarding mean, variance, and skewness?


```{r}
t<- boxplot(calories~type,data=Cereals,main="Cold vs Hot",
              xlab="Type",ylab="calories",
              col="blue")
t

```

Variance: High in cold than hot
Skewness: For Cold, Median is almost near the end of 3 rd quartile, this means that mean is less than median, hence left skewed(Also when we check summary stats we have mean of 107.2 and median of 110 thus indicating left skewed)
For Hot, Mean = Median and No skewness

4) Use a proper visualization to show the distribution of consumer rating by different shelf height. If we
were to predict consumer rating from shelf height, does it appear that we need to keep all three categories
of shelf height or we could combine two of the categories as a new category?

```{r}
d<- boxplot(rating~shelf,data=Cereals,main="cust ratings and shelf link",
              xlab="Ratings",ylab="Shelf",horizontal=TRUE, 
              col="green")
d

```
```{r}
Cereals$shelf<- as.factor(Cereals$shelf)
p <- ggplot(data = Cereals,
             mapping = aes(x = rating, fill = shelf, color = shelf))
p + geom_density(alpha = 0.4)

```
Looking at the box and density plots, 1 and 3 can be combined

5) Get the correlation table for the numerical attributes. In addition, generate a scatter plot array for these
variables. Find one pair of variables with strongest positive correlation and find one pair of variables with
strongest negative correlation

```{r}
library(corrplot)
c<- Cereals[complete.cases(Cereals), ]
c$shelf<- as.numeric(c$shelf)
corrplot(cor(c[,4:16],use="complete.obs"),type="upper",main="Correlation plot")
```

6) Standardize the numerical attributes and then conduct principal component analysis for the numerical variables except consumer rating to derive 4 principal components. Obtain the cumulative proportion of variance (of the numerical variables) captured
by the 4 principal components??

```{r}
scaled.dat <- scale(c[,4:16])
head(scaled.dat)
qs<- c[,4:15]
PCA <- prcomp(qs, center = TRUE, scale = TRUE)
summary(PCA)

```




```{r}
screeplot(PCA, type = "l", npcs = 15, main = "Screeplot of the first 12 PCs")
abline(h = 1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"), col=c("red"), lty=5, cex=0.6)

```

```{r}
var<- PCA$sdev^2 
sum(var[1:4])/sum(var)

```

```{r}
cumpro <- cumsum(PCA$sdev^2 / sum(PCA$sdev^2))
plot(cumpro[0:12], xlab = "PC #", ylab = "Amount of explained variance", main = "Cumulative variance plot")
abline(v = 4, col="blue", lty=5)
abline(h = 0.7308454, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC4"),        col=c("blue"), lty=5, cex=0.6)

```
```{r}
library(factoextra)
fviz_eig(PCA)
```


